A Decision Procedure for (Co)datatypes in SMT Solvers

### 2014-11-19 tex/codatadp.tex
more

  * codatatypes: from a theoretical and implementational point of view, like
      datatypes but:
    * codatatypes are never empty (e.g. finite streams are rejected)
    * no acyclicity (e.g. xs = lcons(0, xs) is satisfiable)
    * instead: bisimilarity
    * (also: ``enumeration'', e.g. of streams)

@@

  * unlike datatypes, codatatypes are never empty

  * how to deal with "wrong" selectors, e.g.
      hd(nil1) = hd(nil2)?
  * ODDITY: Barrett et al. overspecify things (not good w.r.t. SMT-LIB 2, or
    other solvers, e.g., Z3)

>>>>

  * codatatypes: from a theoretical and implementational point of view, like
      datatypes but:
    * codatatypes are never empty (e.g. finite streams are rejected)
    * no acyclicity (e.g. xs = lcons(0, xs) is satisfiable)
    * instead: bisimilarity check
    * (also: ``enumeration'', e.g. of streams)

@@

  * (co)datatype specifications:
    * N mutually recursive types (datatypes or codatatypes -- no mixture)
    * list of n constructors (n >= 1)
    * each constructor has m arguments (m >= 0)
    * each argument has a sort, which may be

  * don't consider mixtures (co)datatypes (cf. Agda), nor polymorphic types, nor
    is it possible to do any nesting; e.g. this is outside our scope:

      datatype tree = Node Int (Array tree)

  * initial algebra
    * ``no junk, no confusion''
    * initial algebra ``exists'' only if well-founded
      otherwise it would be empty, which is disallowed by most formulations of
      first-order logic (including CVC4 and other SMT solvers)
      and higher-order logic (including Isabelle/HOL)

  * final coalgebra
    * this time, no condition: always exists
    e.g. the specification

      codatatype x = X x

    gives rise to a singleton type {X (X (X ...))}.

  * discriminators/selectors for initial algebras
  * constructors for final coalgebras

  * how to deal with "wrong" selectors, e.g.
        hd(nil1) = hd(nil2)?
      * leave them unspecified; hence hd(nil1) = hd(nil2) in some models,
        and not in other models

  * ODDITY: Barrett et al. overspecify things (not good w.r.t. SMT-LIB 2, or
    other solvers, e.g., Z3)

  * axiomatically
    * injectivity
      distinctness
      exhaustiveness
    * datatypes:
      * induction axiom (schema)
        * consequence: no infinite objects (in particular, no cyclic objects)

      * when looking at the universal ground (?) theory:
        enough to consider acyclicity -- no way to specify infinite objects
        otherwise

    * codatatypes:
      * coinduction axiom (schema)
        * consequence: infinite objects are allowed, but
          two objects yielding the same observations must be equal

      * again: when looking at the universal ground (?) theory, enough to
        consider bisimilarity/bisimulation
